Project Description -- Erosion, Transport, Anomalous Diffusion, and Collapse in Granular and Porous Media

### 2019-11-19 docs/proposals/NSF/ProjDesc/ProjDesc.tex
fixed up a few things in proposal

{\bf An ultimate goal of this proposal is to achieve high-fidelity simulations of $\boldsymbol{\mathcal{O}(100-1000)}$ mobile bodies undergoing simultaneous erosion, transport, sedimentation, and compaction, with the computational efficiency required to run large ensembles for statistical analysis.} A major step towards this goal is illustrated in Fig.~\ref{80circ8vel}, which features high-fidelity erosion simulations of an initial configuration of 80 bodies with a volume fraction of $50 \%$. Color indicates the velocity magnitude, which highlights the spontaneous formation of channels during the erosion process. These channels arise from small variations in the initial condition. Once formed, the larger channels transmit more flow which reinforces their growth. The process of channel formation can dramatically alter medium properties and create high levels of anisotropy, which directly impacts dispersion of transported quantities, for example contaminants.

@@

 \todo[inline]{I guess I really took to your strategy of including an example simulation early on in the introduction to draw in the readers interest. So I moved this figure up to the introduction and added some provocative text. Plus there are some visual similarities with the geophysical images in Fig. 1, so it is nice to see them close to one another.}

@@

The PIs co-authored a paper in the {\em Journal of Computational Physics} describing numerical methods for simulating erosion~\cite{Quaife2018} in the Stokes flow regime, which is the regime relevant for the currently proposed groundwater applications. Very recently the PIs, along with postdoctoral scholar Shang-Huan Chiu, submitted a second paper on newly developed methods for precise characterization of transport and dispersion within the special class of porous-media configurations formed by erosion \cite{chi-moo-qua2019}.

@@

%----------------------------------------------------------------------%
\section{Proposed Research} 
%----------------------------------------------------------------------%
\todo[inline]{Discuss what the main research questions are here}
\begin{itemize}[noitemsep]
  \item Characterize anisotropy and permeability to inform homogenized
    models including Darcy's relation.

  \item Improve numerical methods including FMM, preconditioners, and
    quadrature.

  \item Simulate erosion via chemical reactions ie.~dissolution.

  \item Simulate transport of finite Peclet number species in eroded
    geometries.

  \item Understand pore size growth under erosion

  \item New parameterizations using modern machine learning techniques

  \item Validation with experimental results
\end{itemize}

@@

The development of accurate and efficient numerical methods to solve the Stokes equations enables high-fidelity simulations of the erosion process. We will consider erosion induced by either fluid-mechanical forces or by chemical reaction, i.e.~dissolution. In the former case, which we have already begun to investigate~\cite{Quaife2018}, erosion result from the shear stress $\tau = -\left(\nabla \uu + \nabla \uu^T \right)\nn \cdot \ss$, where $\nn$ and $\ss$ are the unit normal and tangent vectors on the surface respectively and $\uu$ is the velocity field as computed from the incompressible Stokes equations~\eqref{eqn:StokesEq}. The motion of the interface is proportional to the absolute value of shear 
\begin{align}
  \Vn = \CE \, \abs{\tau}.
\end{align}
In the case of chemically-induced erosion, a solute concentration field $C$ undergoes advection and diffusion:
\begin{align}
  \label{Ceq}
  \pd{C}{t} + \uu \cdot \grad C = \Diff \Lap C.
\end{align}
Material removal is then governed by Fick's law with interface velocity
\begin{equation}
\label{DissVn}
\Vn = \Diff \pd{C}{\ny}.
\end{equation}
Both erosion processes have been investigated by the PIs and collaborators in the context of single bodies in high Reynolds-number flows~\cite{Ristroph2012, MoorePOF2013, HuangJFM2015, MooreCPAM2017}. In the context of a porous media at low Reynolds-number flows, numerical methods have been developed by the PIs to simulate shear-induced erosion~\cite{Quaife2018, chi-moo-qua2019}. These simulations combine highly accurate boundary integral methods with stable interface evolution techniques.  The boundary integral solver is accelerated by the fast multipole method~\cite{gre-rok1987, gre-gre-may1992} to achieve optimal $O(N)$ complexity. Meanwhile, the so-called {\thL} method is used for interface evolution to prevent distortion and tangling of the interface meshes~\cite{hou1994removing, MoorePOF2013, Quaife2018}.

@@

Figs.~\ref{80circ8vort} and~\ref{80circ8vel} show preliminary results of 80 bodies, initially of random position and size, undergoing mechanical erosion in Stokes flow. This simulation extends previously reported 50-body simulations to 80 bodies~\cite{Quaife2018} due to a recently developed Barycentric interpolation scheme~\cite{bar2014, bar-wu-vee2015, chi-moo-qua2019}, which enables closer contact between bodies and thus more bodies. The color in Fig~\ref{80circ8vort} shows the (normalized) vorticity field surrounding the bodies. Vorticity, since it reduces to shear on solid interfaces, provides a convenient way to visualize local erosion rates, as well as the local flow intensity (Fig.~\ref{80circ8vel}).  Observe that erosion not only reduces the size of solid bodies, but also alters their shapes substantially. The bodies tend to become somewhat polygonal: corners develop connected by relatively flat faces. The number of faces does not appear to be easily predicted, but rather depends on the interaction with neighboring bodies as mediated by the Stokes flow. Notice that, as erosion proceeds, relatively straight channels tend to develop between the bodies. Certain channels that are initially larger can transmit more flow, which promotes local erosion rates and further widens these selected channels. In this way, erosion creates a runaway process in which small differences in initial channel size become amplified. This channelization process is made even more apparent by viewing the velocity magnitude as in Fig.~\ref{80circ8vel}.

@@

Next, we plan to extend the simulation framework to handle chemical dissolution of solid bodies via physical laws~\eqref{Ceq}--\eqref{DissVn}. For this goal, much of the computational infrastructure, such as the boundary-integral solver for the Stokes equations, is already in place, but the coupling to the advection-diffusion equations~\eqref{Ceq} and interface law~\eqref{DissVn} need to be implemented.  We proposed to solve the advection-diffusion equation~\eqref{Ceq} by first using Strang splitting~\cite{str1968} to decompose the PDE into individual PDEs for linear advection and diffusion.  The advection equation will be solved with a semi-Lagrangian method~\cite{rob1981} so that it can be solved on an Eulerian grid without a restrictive CFL condition. PI Quaife has experience with using semi-Lagrangian methods for another type of complex viscous flow~\cite{kab-qua-bir2017}.  The diffusion equation is more challenging to accurately solve in the complex porous geometries.  Two approaches include a memory-intensive space-time heat kernel~\cite{bar-eps-gre-jia-wan2019, jia-gre-wan2015, li-gre2009}, and Rothe's method coupled with a Yukawa equation solver~\cite{kro-qua2010, cau-cho-chr-sea2016}. A particular challenge of the latter method is extending functions from the complex geometry and into a regular geometry, such as a square, so that a high-order volume integral or Fourier method can be applied, and some groups have recently developed new techniques to form such an extension~\cite{fry-kro-tor2019, fry-leh-tor2019}. Alternatively, we propose to use a Laplace Transform to solve the diffusion equation
\begin{alignat}{3}
  \pd{c}{t} &= \frac{1}{\Pe} \Delta c, \quad &&\xx \in \Omega, \: t>0.
\end{alignat}
Applying the Laplace transform, $C(\xx,s) = \mathcal{L}[c](\xx,t)$ satisfies
\begin{align}
  sC(\xx,s) - \frac{1}{\Pe}\Delta C(\xx,s) = c_0(\xx), 
    \quad \xx \in \Omega,
  \label{eqn:DiffusionLaplace}
\end{align}
where $c_0(\xx)$ is the initial concentration. This PDE will be solved with high-order accuracy and efficiently by using fast-multipole-accelerated integral equation methods of the Yukawa equation~\cite{kro-qua2011}, and the concentration will be recovered with an inverse Laplace transform~\cite{jos-war2006}. Once a high-order advection-diffusion solver is in place, by combining it with our integral equation formulation for the Stokes equation and the shear stress, simulations of simultaneous erosion and dissolution, while also including transport and sedimentation, would finally become possible.  Such a framework could accurately simulate the geophysically relevant scenario of karst conduits with embedded granular media. The gradual changes in the medium and karst boundaries due to erosion and dissolution could create conditions favorable for mechanical failure or collapse of the structure, for example sinkholes (see Section~\ref{sec:sinkhole}).

@@

In the infinite Peclet number limit, the concentration is governed by pure advection and the characteristics of the concentration field are identical to the streamlines (Fig.~\ref{fig:100tracers}). The $j^{th}$ trajectory, or characteristic, is governed by $\dot{\ss}_j(t) = \uu(\ss_j(t))$. Once these trajectories are computed, using for example a high-order Runge-Kutta method, a statistical analysis can be used to describe the transport. Two of the most important measurements are the tortuosity and the anomalous dispersion, both which are defined in terms of the length of the trajectories
\begin{align} 
  \lambda_j(t) = \int_{0}^{t} \|\ss'(t')\| dt'.
\end{align}
Upon computing $N_p$ trajectories, the tortuosity is the dimensionless number
\begin{align}
  T = \frac{1}{d} \left(\int_{S} u_1(x_0,y_0)\lambda(y_0) dy_0\, \right)
  \Bigg/ \left(\int_S u_1(x_0,y_0)\, dy_0 \right),
\end{align}
where $S$ is the vertical cross-section at $x=x_0$, $d$ is its length,
$u_1$ is the horizontal velocity, and $\lambda(y_0)$ is the length of
the trajectory starting at $(x_0,y_0)$ and ending at a vertical
cross-section at the other end of the channel. Relationships between the
tortuosity and porosity of a geometry have been developed by other
groups~\cite{kop-kat-tim1996, dud-koz-mat2011, mat-kha-koz2008}, and
preliminary results indicate that such models might be appropriate for
geometries whose porosity is increasing under the process of erosion
(Fig.~\ref{fig:100tortuosity}(a)). However, preliminary results also
indicate that erosion can increase the amount of tortuosity as is
evident towards the at the towards the end of the erosion process in
Fig.~\ref{fig:100tortuosity}(a). Moreover, we have observed a
significant increase in the tortuosity at small porosities
(Fig.~\ref{fig:100tortuosity}(b)).

@@

Upon developing and improving numerical methods to simulate mechanical
and chemical erosion, we plan to develop methods to simulate transport,
gravitational sedimentation, and external buoyant forces of the eroding grains. In addition to the new previously described quadrature, preconditioning, adaptivity, and fast summation methods,  this extension also requires contact forces between bodies to eliminate unphysical contact that is inevitable due to discretization errors. Though conceptually straightforward, implementing contact forces in multibody fluid-structure simulations is far from simple in practice, with several competing possible strategies that are a topic of current research \cite{yan2019computing}.

\todo[inline]{Bryan, do you know any other references to include after `topic of current research'? I included Shelley and Yan's method. There were a bunch of others at Oaxaca but I don't remember names.}

@@

%----------------------------------------------------------------------%
\section{Timeline and Concluding Remarks}

>>>>

{\bf An ultimate goal of this proposal is to achieve high-fidelity simulations of $\boldsymbol{\mathcal{O}(10^2-10^3)}$ mobile bodies undergoing simultaneous erosion, transport, sedimentation, and compaction, with the computational efficiency required to run large ensembles for statistical analysis.} A major step towards this goal is illustrated in Fig.~\ref{80circ8vel}, which features high-fidelity erosion simulations of an initial configuration of 80 bodies with a volume fraction of $50 \%$. Color indicates the velocity magnitude, which highlights the spontaneous formation of channels during the erosion process. These channels arise from small variations in the initial condition. Once formed, the larger channels transmit more flow which reinforces their growth. The process of channel formation can dramatically alter medium properties and create high levels of anisotropy, which directly impacts dispersion of transported quantities, for example contaminants.

@@

@@

The PIs co-authored a paper in the {\em Journal of Computational Physics} describing numerical methods for simulating erosion in the Stokes flow regime~\cite{Quaife2018}, which is the regime relevant for the currently proposed groundwater applications. Very recently the PIs, along with postdoctoral scholar Shang-Huan Chiu, submitted a second paper on newly developed methods for precise characterization of transport and dispersion within the special class of porous-media configurations formed by erosion \cite{chi-moo-qua2019}.

@@

%----------------------------------------------------------------------%
\section{Proposed Research} 
%----------------------------------------------------------------------%
The goal of the proposed research is to develop new numerical methods to
simulate erosion processes. Specific research tasks will include:
\begin{itemize}[noitemsep]
  \item Simulating chemical erosion.

  \item Characterizing anisotropy and permeability to inform
    homogenized models such as Darcy's relation.

  \item Simulating the transport of finite Peclet number concentrations
    inside eroded porous geometries.

  \item Improving numerical methods including fast summation methods,
    preconditioners, and quadrature.

  \item Developing new models to characterize the effect of erosion on
    pore size dynamics.

  \item Developing new parameterizations using modern machine learning
    techniques.

  \item Validating simulations with in-house experimental results.

  \item \todo[inline]{Something about transport, sedimentation, etc.}
\end{itemize}

@@

The development of accurate and efficient numerical methods to solve the Stokes equations enables high-fidelity simulations of the erosion process. We will consider erosion induced by either fluid-mechanical forces or by chemical reaction, i.e.~dissolution. In the former case, which we have already begun to investigate~\cite{Quaife2018}, erosion result from the shear stress $\tau = -\left(\nabla \uu + \nabla \uu^T \right)\nn \cdot \ss$, where $\nn$ and $\ss$ are the unit normal and tangent vectors on the surface respectively and $\uu$ is the velocity field as computed from the incompressible Stokes equations~\eqref{eqn:StokesEq}. The motion of the interface is proportional to the absolute value of shear 
\begin{align}
  \Vn = \CE \, \abs{\tau}.
\end{align}
In the case of chemically-induced erosion, a solute concentration field $C$ undergoes advection and diffusion:
\begin{align}
  \label{Ceq}
  \pd{C}{t} + \uu \cdot \grad C = \Diff \Lap C.
\end{align}
Material removal is then governed by Fick's law with interface velocity
\begin{equation}
\label{DissVn}
\Vn = \Diff \pd{C}{\ny}.
\end{equation}
Both erosion processes have been investigated by the PIs and collaborators in the context of single bodies in high Reynolds-number flows~\cite{Ristroph2012, MoorePOF2013, HuangJFM2015, MooreCPAM2017}. In the context of a porous media at low Reynolds-number flows, numerical methods have been developed by the PIs to simulate shear-induced erosion~\cite{Quaife2018, chi-moo-qua2019}. These simulations combine highly accurate boundary integral methods with stable interface evolution techniques.  The boundary integral solver is accelerated by the fast multipole method~\cite{gre-rok1987, gre-gre-may1992} to achieve optimal $\mathcal{O}(N)$ complexity. Meanwhile, the so-called {\thL} method is used for interface evolution to prevent distortion and tangling of the interface meshes~\cite{hou1994removing, MoorePOF2013, Quaife2018}.

@@

Figs.~\ref{80circ8vel} and~\ref{80circ8vort} show preliminary results of 80 bodies, initially of random position and size, undergoing mechanical erosion in Stokes flow. This simulation extends previously reported 50-body simulations to 80 bodies~\cite{Quaife2018} due to a recently developed Barycentric interpolation scheme~\cite{bar2014, bar-wu-vee2015, chi-moo-qua2019}, which enables closer contact between bodies and thus more bodies. The color in Fig~\ref{80circ8vort} shows the (normalized) vorticity field surrounding the bodies. Vorticity, since it reduces to shear on solid interfaces, provides a convenient way to visualize local erosion rates, as well as the local flow intensity (Fig.~\ref{80circ8vel}).  Observe that erosion not only reduces the size of solid bodies, but also alters their shapes substantially. The bodies tend to become somewhat polygonal: corners develop connected by relatively flat faces. The number of faces does not appear to be easily predicted, but rather depends on the interaction with neighboring bodies as mediated by the Stokes flow. Notice that, as erosion proceeds, relatively straight channels tend to develop between the bodies. Certain channels that are initially larger can transmit more flow, which promotes local erosion rates and further widens these selected channels. In this way, erosion creates a runaway process in which small differences in initial channel size become amplified. This channelization process is made even more apparent by viewing the velocity magnitude as in Fig.~\ref{80circ8vel}.

@@

Next, we plan to extend the simulation framework to handle chemical dissolution of solid bodies via physical laws~\eqref{Ceq}--\eqref{DissVn}. For this goal, much of the computational infrastructure, such as the boundary-integral solver for the Stokes equations, is already in place, but the coupling to the advection-diffusion equations~\eqref{Ceq} and interface law~\eqref{DissVn} need to be implemented.  We proposed to solve the advection-diffusion equation~\eqref{Ceq} by first using Strang splitting~\cite{str1968} to decompose the PDE into individual PDEs for linear advection and diffusion.  The advection equation will be solved with a semi-Lagrangian method~\cite{rob1981} so that it can be solved on an Eulerian grid without a restrictive CFL condition. PI Quaife has experience with using semi-Lagrangian methods for another type of complex viscous flow~\cite{kab-qua-bir2017}.  The diffusion equation is more challenging to accurately solve in the complex porous geometries.  Two approaches include a memory-intensive space-time heat kernel~\cite{bar-eps-gre-jia-wan2019, jia-gre-wan2015, li-gre2009}, and Rothe's method coupled with a Yukawa equation solver~\cite{kro-qua2010, cau-cho-chr-sea2016}. A particular challenge of the latter method is extending functions from the complex geometry and into a regular geometry, such as a square, so that a high-order volume integral or Fourier method can be applied, and some groups have recently developed new techniques to form such an extension~\cite{fry-kro-tor2019, fry-leh-tor2019}. Alternatively, we propose to use a Laplace Transform to solve the diffusion equation
\begin{alignat}{3}
  \pd{c}{t} &= \frac{1}{\Pe} \Delta c, \quad &&\xx \in \Omega, \: t>0.
\end{alignat}
Applying the Laplace transform, $C(\xx,s) = \mathcal{L}[c](\xx,t)$ satisfies
\begin{align}
  sC(\xx,s) - \frac{1}{\Pe}\Delta C(\xx,s) = c_0(\xx), 
    \quad \xx \in \Omega, \: s \in \mathbb{C},
  \label{eqn:DiffusionLaplace}
\end{align}
where $c_0(\xx)$ is the initial concentration. This PDE will be solved with high-order accuracy and efficiently by using fast-multipole-accelerated integral equation methods of the Yukawa equation~\cite{kro-qua2011}, and the concentration will be recovered with an inverse Laplace transform~\cite{jos-war2006}. Once a high-order advection-diffusion solver is in place, by combining it with our integral equation formulation for the Stokes equation and the shear stress, simulations of simultaneous erosion and dissolution, while also including transport and sedimentation, would finally become possible.  Such a framework could accurately simulate the geophysically relevant scenario of karst conduits with embedded granular media. The gradual changes in the medium and karst boundaries due to erosion and dissolution could create conditions favorable for mechanical failure or collapse of the structure, for example sinkholes (see Section~\ref{sec:sinkhole}).

@@

In the infinite Peclet number limit, the concentration is governed by pure advection and the characteristics of the concentration field are identical to the streamlines (Fig.~\ref{fig:100tracers}). The $j^{th}$ trajectory, or characteristic, is governed by $\dot{\ss}_j(t) = \uu(\ss_j(t))$. Once these trajectories are computed, using for example a high-order Runge-Kutta method, a statistical analysis can be used to describe the transport. Two of the most important measurements are the tortuosity and the anomalous dispersion, both which are defined in terms of the length of the trajectories. Upon computing $N_p$ trajectories, the tortuosity is the dimensionless number
\begin{align}
  T = \frac{1}{d} \left(\int_{S} u_1(x_0,y_0)\lambda(y_0) dy_0\, \right)
  \Bigg/ \left(\int_S u_1(x_0,y_0)\, dy_0 \right),
\end{align}
where $S$ is the vertical cross-section at $x=x_0$, $d$ is its length, $u_1$ is the horizontal velocity, and $\lambda(y_0)$ is the length of the trajectory starting at $(x_0,y_0)$ and ending at a vertical cross-section at the other end of the channel. Relationships between the tortuosity and porosity of a geometry have been developed by other groups~\cite{kop-kat-tim1996, dud-koz-mat2011, mat-kha-koz2008}, and preliminary results indicate that such models might be appropriate for geometries whose porosity is increasing under the process of erosion (Fig.~\ref{fig:100tortuosity}(a)). However, preliminary results also indicate that erosion can increase the amount of tortuosity as is evident towards the at the towards the end of the erosion process in Fig.~\ref{fig:100tortuosity}(a). Moreover, we have observed a significant increase in the tortuosity at small porosities (Fig.~\ref{fig:100tortuosity}(b)).

@@

Upon developing and improving numerical methods to simulate mechanical and chemical erosion, we plan to develop methods to simulate transport, gravitational sedimentation, and external buoyant forces of the eroding grains. In addition to the new previously described quadrature, preconditioning, adaptivity, and fast summation methods,  this extension also requires contact forces between bodies to eliminate unphysical contact that is inevitable due to discretization errors. Though conceptually straightforward, implementing contact forces in multibody fluid-structure simulations is far from simple in practice, with several competing possible strategies that are a topic of current research \cite{yan2019computing, lu-rah-zor2017, bys-sha-qua2019}.

@@

%----------------------------------------------------------------------%
\section{Timeline}
